---
title: "Homework 2"
author: "Ankur Patel"
date: "9/16/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(dplyr)
library(tidyverse)
```

## Problem 3:
In my future work, I think version control will be useful mostly in collaborative settings because it will enable my team members and myself to keep track of each other's changes. However, if I am the sole person working on a script I may not make use of it as I will keep track of changes through comments directly on the script. I think that it is mostly useful when a script is being worked on by at least 2 people and is a good safeguard against fatal coding changes.

## Problem 4:

(a) We are looking at the sensory data from five operators from the Wu and Hamada book. First, we will get the raw data using the URL.

```{r sensory_data, echo=TRUE, cache=FALSE, include=TRUE}
#Code to get "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat"
sensory_url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/Sensory.dat"
sensory_data_raw <- fread(sensory_url,skip = 2,data.table = FALSE,header = FALSE, fill = TRUE)
```
Now we will tidy the sensory data using base R.
```{r sensory_data_tidy_baseR,cache = FALSE, echo = TRUE, include = TRUE}
#create an items_column which will be on the left side of the dataframe
items_column <- vector(length = 0)
for (i in 1:10)
{
  items_column <- append(items_column, rep(i,3))
}
#the item numbers are woven into the first column so we replace them with NA's #using the two lines of code below and call the new column first col_clean
first_col_clean <- sensory_data_raw$V1
first_col_clean[seq(1,30,3)] <- NA
sensory_data_tidy_baseR <- cbind(first_col_clean,sensory_data_raw[,2:6])
#we store the rows to fix in rows_to_fix
rows_to_fix <- seq(1,30,3)
#the for loop below shifts the rows to fix one entry to the left and the last entry is NA, ensuring that the operator and values are now aligned
for (i in 1:length(rows_to_fix))
{
  curr_row <- rows_to_fix[i]
  cleaned_row <- c(sensory_data_tidy_baseR[curr_row,2:6],NA)
  sensory_data_tidy_baseR[curr_row,] <- cleaned_row
}
#rename the first column
names(sensory_data_tidy_baseR)[1] <- "V1"
#drop the last column since it is all NA
sensory_data_tidy_baseR <- sensory_data_tidy_baseR[,1:5]
#bind the items_column and rename the columns
sensory_data_tidy_baseR <- cbind(items_column,sensory_data_tidy_baseR)
names(sensory_data_tidy_baseR)<- c("Item","1","2","3","4","5")
#the current form of sensory_data_tidy_baseR is what we will use for the tidyverse version so we make a copy of it here
sensory_data_tidy_tidyverse <- copy(sensory_data_tidy_baseR)
#operator_column repeats the operator values in a sequence 1,...,5 30 times which will be one of the columns of the final 150 x 3 dataframe
operator_column <- rep(seq(1,5),30)
#the measurement column turns the  data in the rows into a vector
measurement_column <- vector(length = 0)
for (i in 1:dim(sensory_data_tidy_baseR)[1])
{
  #unlist is used to turn the rows into vectors
  measurement_column <- append(measurement_column,unlist(sensory_data_tidy_baseR[i,2:6],use.names = FALSE))
}
#the final items column should have 150 entries
items_column_final <- vector(length = 0)
for (i in 1:10)
{
  items_column_final <- append(items_column_final, rep(i,15))
}
sensory_data_tidy_baseR <- data.frame(cbind(items_column_final,operator_column,measurement_column))
names(sensory_data_tidy_baseR) <-c("Item", "Operator","Data")
print(head(sensory_data_tidy_baseR))

```
Now we will tidy the sensory data using tidyverse
```{r sensory_data_tidy_tidyverse,cache = FALSE, echo = TRUE, include = TRUE}
sensory_data_tidy_tidyverse <- sensory_data_tidy_tidyverse %>% gather(key = "Operator", value = "Data", 2,3,4,5,6)
print(head(sensory_data_tidy_tidyverse))
```
Now we will do a summary and plot of the sensory data.
```{r sensory_data_summary_plot,cache = FALSE, echo = TRUE, include = TRUE}
knitr::kable(summary(sensory_data_tidy_tidyverse))
boxplot(sensory_data_tidy_tidyverse$Data~sensory_data_tidy_tidyverse$Item)

```
(b) Now we will look at the olympic data. First we will read in the raw data from the URL.
```{r olympic_data,cache = FALSE, echo = TRUE, include = TRUE}
#Assign the url to a variable
olympic_url <- "http://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/LongJumpData.dat"
#Read in the raw data
olympic_data_raw <- fread(olympic_url,skip = 1,data.table = FALSE,header = FALSE, fill = TRUE)

```
Next we will tidy the olympic data using base R.
```{r olympic_data_tidy_baseR,cache = FALSE, echo = TRUE, include = TRUE}
#years_columns are the positions of the columns that have year values
years_columns <- c(1,3,5,7)
#jump_columns are the positions of the columns that have values for the long jump
jump_columns <- c(2,4,6,8)
#standard_year is the standardized year (by adding 1900) of the stacked Year columns using years_columns above
standard_year <- 1900+stack(olympic_data_raw[,years_columns])[1]
stacked_jump <- stack(olympic_data_raw[,jump_columns])[1]
olympic_data_tidy_baseR <- data.frame(cbind(standard_year, stacked_jump))
names(olympic_data_tidy_baseR) <- c("Year", "Long Jump")
print(head(olympic_data_tidy_baseR))
```
Next we will tidy the olympic data using tidyverse
```{r olympic_data_tidy_tidyverse,cache = FALSE, echo = TRUE, include = TRUE}
#use dplyr select to get the year and long jump columns separately into dataframes
years_frame <-  select(olympic_data_raw,1,3,5,7)
jump_frame <-  select(olympic_data_raw,2,4,6,8)
#stack the dataframes
years_stacked <- 1900+stack(years_frame)[1]
jump_stacked <- stack(jump_frame)[1]
#bind the stacked dataframes and restore the names
olympic_data_tidy_tidyverse <- data.frame(cbind(years_stacked,jump_stacked))
names(olympic_data_tidy_tidyverse) <- c("Year","Long Jump")
print(head(olympic_data_tidy_tidyverse))
```
Finally we create a summary and plot of the olympic data
```{r olympic_summary_plot,cache = FALSE, echo = TRUE, include = TRUE}
knitr::kable(summary(olympic_data_tidy_tidyverse))
plot(olympic_data_tidy_tidyverse$Year,olympic_data_tidy_tidyverse$`Long Jump`)
```
(c) Now we will look at the brain and body weight data for various species. First we will import the raw data from the URL.
```{r species_data,cache = FALSE, echo = TRUE, include = TRUE}
#Specify and assign the url to a variable
species_url <- "https://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/BrainandBodyWeight.dat"
#Read in the raw data
species_data_raw <- fread(species_url,skip = 1,data.table = FALSE,header = FALSE, fill = TRUE)
```
Next we will tidy the species data using base R.
```{r species_data_tidy_baseR,cache = FALSE, echo = TRUE, include = TRUE}
#body_columns and brain_columns are the position of the columns in the raw species data for body weight and brain weight respectively
body_columns <- seq(1,6,2)
brain_columns <- seq(2,6,2)
#body_vector and brain_vector are the stacked vectors of body and brain weight
#extracted from species_data_raw by using body_columns and brain_columns
body_vector <- stack(species_data_raw[,body_columns])[1]
brain_vector <- stack(species_data_raw[,brain_columns])[1]
#the id corresponding to a unique species
species_id <- seq(1,62)
#Now we combine the stacked vectors (the first 62 entries since there are only 62 species) into a data frame and add the species id column 
species_data_tidy_baseR<-data.frame(cbind(species_id,body_vector[1:62,1],brain_vector[1:62,1]))
names(species_data_tidy_baseR) <- c("Species ID","Body Weight(Kg)", "Brain Weight(g)")
print(head(species_data_tidy_baseR))
```
Now we will tidy the species data using tidyverse
```{r species_data_tidy_tidyverse,cache = FALSE, echo = TRUE, include = TRUE}
#use dplyr to select the brain and body weight columns into their own data frame
#brain_frame and body_frame contain the brain and body weight columns 
brain_frame <- select(species_data_raw,seq(2,6,2))
body_frame <- select(species_data_raw,seq(1,6,2))
#stack the dataframes
brain_stack <- stack(brain_frame)
body_stack <- stack(body_frame)
#build the tidy dataframe using data.frame and cbind from the stacked dataframes and restore the original names
species_data_tidy_tidyverse <- data.frame(cbind(species_id,body_vector[1:62,1],brain_vector[1:62,1]))
names(species_data_tidy_tidyverse) <- c("Species ID", "Body Weight (Kg)", "Brain Weight (g)")
print(head(species_data_tidy_tidyverse))

```
Now we will do a summary and plot of the species data
```{r species_data_summary_plot,cache = FALSE, echo = TRUE, include = TRUE }
knitr::kable(summary(species_data_tidy_tidyverse[,2:3]))
plot(species_data_tidy_tidyverse$`Body Weight (Kg)`,species_data_tidy_tidyverse$`Brain Weight (g)`)
```
(d) Now we will look at the tomato yield data. We will first assign the URL to a variable and then import the raw data.
```{r tomato_data,cache = FALSE, echo = TRUE, include = TRUE}
tomato_url <- "https://www2.isye.gatech.edu/~jeffwu/wuhamadabook/data/tomato.dat"
#Read in the raw data
tomato_data_raw <- fread(tomato_url,skip = 1,data.table = FALSE,header = TRUE, fill = TRUE)
```
Now we will tidy the data using base R.
```{r tomato_data_tidy_baseR,cache = FALSE, echo = TRUE, include = TRUE}
#make a copy of the raw data
tomato_data_tidy_baseR <- copy(tomato_data_raw)
#rename the columns and apply it to the copy
names(tomato_data_tidy_baseR) <- c("Variety", "10000", "20000", "30000")
#there are three measurements for each of the planting densities ("10000","20000","30000"). So for each variety there are 9 measurements and we will store them in Ife_yield and PusaEarlyDwarf respectively.
Ife_yield <- vector(length = 0)
PusaEarlyDwarf <- vector(length = 0)
#populate Ife_yield and PusaEarlyDwarf by unlisting, string splitting and converting to numeric the data in the rows
for (j in 2:4)
{
  Ife_yield <- append(Ife_yield,as.numeric(unlist(strsplit(tomato_data_tidy_baseR[1,j], split = ","))))
  PusaEarlyDwarf <- append(PusaEarlyDwarf,as.numeric(unlist(strsplit(tomato_data_tidy_baseR[2,j], split = ","))))
}
measurements <- c(Ife_yield, PusaEarlyDwarf)
#set up the column for planting density
plant_density <- c(rep(10000,3),rep(20000,3),rep(30000,3))
plant_density <- rep(plant_density,2)
#set up the column for variety names
variety_names <- c(rep(tomato_data_tidy_baseR[1,1],9), rep(tomato_data_tidy_baseR[2,1],9))
#construct the dataframe
tomato_data_tidy_baseR <-  data.frame(cbind(variety_names,plant_density,measurements))
#give appropriate names
names(tomato_data_tidy_baseR) <- c("Variety", "Planting Density", "Measurement")
print(head(tomato_data_tidy_baseR))
```
Now we will tidy the tomato data using tidyverse
```{r tomato_data_tidy_tidyverse,cache = FALSE, echo = TRUE, include = TRUE}
#copy the raw data as tomato+data_tidy_tidyverse and reassign column names
tomato_data_tidy_tidyverse <- tomato_data_raw
names(tomato_data_tidy_tidyverse) <- c("Variety", "10000", "20000", "30000")
#densityi_col corresponds to column i+1. So for example density1_col corresponds to the column header 10000.
density1_col <- c(as.numeric(unlist(strsplit(tomato_data_tidy_tidyverse[1,2], split = ","))),as.numeric(unlist(strsplit(tomato_data_tidy_tidyverse[2,2], split = ","))))
density2_col <- c(as.numeric(unlist(strsplit(tomato_data_tidy_tidyverse[1,3], split = ","))),as.numeric(unlist(strsplit(tomato_data_tidy_tidyverse[2,3], split = ","))))
density3_col <- c(as.numeric(unlist(strsplit(tomato_data_tidy_tidyverse[1,4], split = ","))),as.numeric(unlist(strsplit(tomato_data_tidy_tidyverse[2,4], split = ","))))
#variety_col will list the variety names for each measurement
variety_col <- c(rep(tomato_data_tidy_tidyverse[1,1],3),rep(tomato_data_tidy_tidyverse[2,1],3))
#get the untidy dataframe ready for using the gather function
tomato_data_tidy_tidyverse <- data.frame(cbind(variety_col,density1_col,density2_col,density3_col))
names(tomato_data_tidy_tidyverse) <- c("Variety", "10000", "20000", "30000")
#use the tidyverse gather function to get the final tidy dataframe
tomato_data_tidy_tidyverse <- tomato_data_tidy_tidyverse %>% gather(key = "Planting Density", value = "Measurement",2,3,4)
tomato_data_tidy_tidyverse$`Planting Density` <- as.integer(tomato_data_tidy_tidyverse$`Planting Density`)
tomato_data_tidy_tidyverse$Measurement <- as.numeric(tomato_data_tidy_tidyverse$Measurement)
print(head(tomato_data_tidy_tidyverse))
```
Now we will do a summary and plot of the tomato yield data
```{r tomato_data_summary_plot,cache = FALSE, echo = TRUE, include = TRUE}
#gives a summary table and a boxplot of measurement by variety
knitr::kable(summary(tomato_data_tidy_tidyverse[2:3]))
boxplot(tomato_data_tidy_tidyverse$Measurement ~ tomato_data_tidy_tidyverse$Variety)
```